---
title: "Equine Colic data-set"
output: html_document
---

For details see (https://archive.ics.uci.edu/ml/datasets/Horse+Colic)

To load into R use...

```{r, eval = FALSE}
library(tidyverse)

colic <- read_csv("colic_data.csv")
```




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE}
library(tidyverse)
```

```{r, include = FALSE}
# parse names text - we'll remove columns 25, 26, & 27 from data as lesion and hard to parse

colic_names <- read_lines("horse-colic.names.txt") %>%
  keep(~ str_detect(.x, "[0-9]:")) %>%
  str_remove(" *[0-9]+:") %>%
  discard(~ str_detect(.x, "25")) %>%
  trimws()
```

```{r, include = FALSE}
colic <- read_delim(
  "horse-colic.data.txt",
  delim = " ",
  col_names = FALSE,
  na = "?"
) %>%
  select(-(25:27)) %>%
  set_names(colic_names) %>%
  janitor::clean_names() %>%
  rename(pain = pain_a_subjective_judgement_of_the_horses_pain_level) %>%
  mutate(
    surgery = fct_recode(
      factor(surgery),
      "yes" = "1",
      "no" = "2"
    ),
    age = fct_recode(
      factor(age),
      "adult" = "1",
      "young" = "9"
    ),
    temperature_of_extremities = fct_recode(
      factor(temperature_of_extremities),
      "normal" = "1",
      "warm" = "2",
      "cool" = "3",
      "cold" = "4"
    ),
    peripheral_pulse = fct_recode(
      factor(peripheral_pulse),
      "normal" = "1",
      "increased" = "2",
      "reduced" = "3",
      "absent" = "4"
    ),
    pain = fct_recode(
      factor(pain),
      "no pain" = "1",
      "depressed" = "2",
      "intermittent mild pain" = "3",
      "intermittent severe pain" = "4",
      "continuous severe pain" = "5"
    ),
    abdominal_distension = fct_recode(factor(abdominal_distension),
      "none" = "1",
      "slight" = "2",
      "moderate" = "3",
      "severe" = "4"
    ),
    rectal_examination_feces = fct_recode(factor(rectal_examination_feces),
      "normal" = "1",
      "increased" = "2",
      "decreased" = "3",
      "absent feces" = "4"
    ),
    outcome = fct_recode(factor(outcome),
      "lived" = "1",
      "died_or_euth" = "2",
      "died_or_euth" = "3"
    ),
    surgical_lesion = fct_recode(factor(surgical_lesion),
      "yes" = "1",
      "no" = "2"
    ),
    capillary_refill_time = fct_recode(factor(capillary_refill_time),
      "< 3 seconds" = "1",
      ">= 3 seconds" = "2"
    ),
    peristalsis = fct_recode(factor(peristalsis),
      "hypermotile" = "1",
      "normal" = "2",
      "hypomotile" = "3",
      "absent" = "4"
    )
  ) %>%
  select(
    -cp_data, -abdominocentesis_appearance, -abdomcentesis_total_protein,
    -nasogastric_tube, -nasogastric_reflux, -nasogastric_reflux_ph,
    -abdomen, -hospital_number, -mucous_membranes
  ) %>%
  filter(!is.na(outcome)) %>%
  mutate(total_protein = case_when(
    total_protein > 25 ~ total_protein / 10,
    TRUE ~ total_protein
  )) %>%
  write_csv("colic_data.csv")
```

Creators: 
Mary McLeish & Matt Cecile 
Department of Computer Science 
University of Guelph 
Guelph, Ontario, Canada N1G 2W1 
mdmcleish '@' water.waterloo.edu 

Donor: 
Will Taylor (taylor '@' pluto.arc.nasa.gov)

Data Set Information:

Many Data types: (continuous, discrete, and nominal)

Attribute Information:

1: surgery
1 = Yes, it had surgery 
2 = It was treated without surgery 

2: age 
1 = Adult horse 
2 = Young (< 6 months) 

4: rectal_temperature 
- linear 
- in degrees celsius. 
- An elevated temp may occur due to infection. 
- temperature may be reduced when the animal is in late shock 
- normal temp is 37.8 
- this parameter will usually change as the problem progresses, eg. may start out normal, then become elevated because of the lesion, passing back through the normal range as the horse goes into shock 

5: pulse 
- linear 
- the heart rate in beats per minute 
- is a reflection of the heart condition: 30 -40 is normal for adults 
- rare to have a lower than normal rate although athletic horses may have a rate of 20-25 
- animals with painful lesions or suffering from circulatory shock may have an elevated heart rate 

6: respiratory_rate 
- linear 
- normal rate is 8 to 10 
- usefulness is doubtful due to the great fluctuations 

7: temperature_of_extremities 
- a subjective indication of peripheral circulation 
- possible values: 
1 = Normal 
2 = Warm 
3 = Cool 
4 = Cold 
- cool to cold extremities indicate possible shock 
- hot extremities should correlate with an elevated rectal temp. 

8: peripheral_pulse 
- subjective 
- possible values are: 
1 = normal 
2 = increased 
3 = reduced 
4 = absent 
- normal or increased p.p. are indicative of adequate circulation while reduced or absent indicate poor perfusion 

10: capillary_refill_time 
- a clinical judgement. The longer the refill, the poorer the circulation 
- possible values 
1 = < 3 seconds 
2 = >= 3 seconds 

11: pain
- a subjective judgement of the horse's pain level 
- possible values: 
1 = alert, no pain 
2 = depressed 
3 = intermittent mild pain 
4 = intermittent severe pain 
5 = continuous severe pain 
- should NOT be treated as a ordered or discrete variable! 
- In general, the more painful, the more likely it is to require surgery 
- prior treatment of pain may mask the pain level to some extent 

12: peristalsis 
- an indication of the activity in the horse's gut. As the gut becomes more distended or the horse becomes more toxic, the activity decreases 
- possible values: 
1 = hypermotile 
2 = normal 
3 = hypomotile 
4 = absent 

13: abdominal_distension 
- An IMPORTANT parameter. 
- possible values 
1 = none 
2 = slight 
3 = moderate 
4 = severe 
- an animal with abdominal distension is likely to be painful and have reduced gut motility. 
- a horse with severe abdominal distension is likely to require surgery just tio relieve the pressure 

17: rectal_examination
- feces 
- possible values 
1 = normal 
2 = increased 
3 = decreased 
4 = absent 
- absent feces probably indicates an obstruction 

19: packed_cell_volume 
- linear 
- the # of red cells by volume in the blood 
- normal range is 30 to 50. The level rises as the circulation becomes compromised or as the animal becomes dehydrated. 

20: total_protein 
- linear 
- normal values lie in the 6-7.5 (gms/dL) range 
- the higher the value the greater the dehydration 

23: outcome 
- what eventually happened to the horse? 
- possible values: 
1 = lived 
2 = died or euthanized


24: surgical_lesion 
- retrospectively, was the problem (lesion) surgical? 
- all cases are either operated upon or autopsied so that this value and the lesion type are always known 
- possible values: 
1 = Yes 
2 = No 




